Liveness-Based Garbage Collection for Lazy Languages

### 2016-04-27 paper-for-archive/paper.tex
Updated arXiv version with the changes

\documentclass[9pt,preprint,nonatbib]{sigplanconf}
\usepackage{balance}
\usepackage{float}
\usepackage{mathrsfs}
%%\usepackage{lineno}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{etex}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsmath}
%\usepackage[pdf]{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
%\usepackage{pst-rel-points}
\usepackage{bcprules}
\usepackage{subfigure}
\usepackage[boxed]{algorithm2e}
\usepackage{makecell}

% to save space in lists
\usepackage{enumitem}
\setlist{nolistsep}

@@

% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.
%\cmt{{
%
%\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors
%section.
%
%\author{
% 1st. author
\authorinfo {Prasanna Kumar. K} 
       {IIT Bombay, Mumbai 400076, India}
       {prasannak@cse.iitb.ac.in}
% 2nd. author
\authorinfo {Amitabha Sanyal} 
       {IIT Bombay, Mumbai 400076, India}
       {as@cse.iitb.ac.in}
% 3rd. author
\authorinfo {Amey Karkare}
       {IIT Kanpur, Kanpur 208016, India}
       {karkare@cse.iitk.ac.in}

@@

While  the runtime  system  of most  functional  languages includes  a
garbage collector to efficiently  reclaim memory, empirical studies on
Scheme~\cite{karkare06effectiveness}   and,   more   importantly,   on
Haskell~\cite{rojemo96lag} programs have shown that garbage collectors
leave uncollected a large number  of memory objects that are reachable
but not live (here {\em live} means the object can potentially be used
by the program at a  later stage).  This results in unnecessary memory
retention.

@@

\added{While  the idea  of  using static  analysis  to improve  memory
  utilization  has  been  shown  to   be  effective  for  {\em  eager}
  languages~\cite{asati14lgc,       HofmannJ03,       inoue88analysis,
    lee05static}, a straightforward extension  of the technique is not
  possible  for  lazy  languages,  where  heap-allocated  objects  may
  include   closures    (runtime   representations    of   unevaluated
  expressions).   }   \todelete{Designing  a  liveness  based  garbage
  collector   for  lazy   languages  poses   significant  challenges.}
Firstly, since  data is made  live by  evaluation of closures,  and in
lazy languages  the place in  the program where this  evaluation takes
place cannot  be statically determined, laziness  complicates liveness
analysis itself.  Moreover, for liveness-based  GC to be effective, we
need to  extend it  to closures  apart from  evaluated data.   Since a
closure can escape the scope in which  it was created, it has to carry
the  liveness  information  of   its  free  variables.   As  execution
progresses  and possible  future uses  are eliminated,  we update  the
liveness  information  in  a  closure with  a  more  precise  version.
Because  of   these  reasons,  the  garbage   collector  also  becomes
significantly more  complicated than a liveness-based  collector for a
eager language.

@@

  Figure~\ref{fig:mot-example}   shows  an   example
program and the state of the heap at the program point $\pi$\footnote{We write
$\pi\!:\!e$ to  associate a  label $\pi$ with  the program  point just
before  the  expression  $e$. The label is not part of the syntax.},  i.e., just  before  the  evaluation  of
$(\length\ \pz)$.  The heap is represented  by a graph in which a node
either  represents  atomic  values   ($\NIL$,  integers  etc.),  or  a
\CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
(represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
  references} and represent values of variables or fields.  The figure
shows the  lists \px\  and \pz\  partially evaluated  due to  the
\SIF\ condition 
(\NULLQ~(\CAR~\pz)).
The edges  shown by thick  arrows are those  which are live  at $\pi$.
%% Only those  cell which have a future use  should be preserved
%% during garbage collection, all other
%% cells can be  reclaimed.
Thus if a GC takes  place at $\pi$ with the heap shown
in Figure~\ref{fig:mot-example}(b),  a liveness-based  collector (LGC)
will preserve only  the cells referenced by $\pz$, and  the live cells
constituting the  closure referenced by $(\CDR~\pz)$.   In contrast, a
reachability-based collector (RGC) will preserve all cells.

In  this work  we show  that  static analysis  of heap  data can  help
garbage  collectors   in  reclaiming   more  garbage.    The  specific
contributions of this paper are:
\added{
\begin{itemize}
\item We  propose a liveness  based GC scheme  for a lazy  first order
  functional language and  prove its correctness.  To the  best of our
  knowledge,  this is  the first  work on  liveness-based GC  for lazy
  languages.       All      previous      attempts~\cite{shaham01heap,
    ran.shaham-sas03,          shaham02estimating,         asati14lgc,
    karkare06effectiveness} have  involved either imperative  or eager
  functional languages.  The novelty of our approach is in handling of
  liveness  of closures,  an  issue  that arises  due  to laziness  at
  runtime.

@@

\begin{figure}[t]\footnotesize
\renewcommand{\arraystretch}{0.9}
\begin{eqnarray*}
   p \in \mathit{Prog} & \!\!\!::=\!\!\! & d_1 \ldots d_n \,\,\,\, e_\mainpgm
   \hspace{5em} \,\,\,\,\,\,\,\,\,\; \mbox{\em --- program}\\
    \mathit{df} \in Fdef & \!\!\!::=\!\!\! & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots
\,\,x_n)\,\,
    e)
    \hspace{0.2em} \ \ \ \ \ \ \ \ \  \mbox{\em --- function def} \\
e \in \mathit{Expr} & \!\!\!::=\!\!\! &
\left\{\begin{array}{@{}ll@{\hspace{1em}}l}
       (\SIF\,\, x\,\, e_1\,\, e_2) && \!\!\!\mbox{\em --- conditional} \\
       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) &&\!\!\! \mbox{\em --- let
binding} \\
       (\SRETURN\,\, x) && \!\!\!\mbox{\em --- return from function}
    \end{array}\right. \\
s \in \mathit{App} & \!\!\!::=\!\!\!  &
\left\{\begin{array}{@{}l@{\hspace{1.2em}}l}
       k & \mbox{\em --- constant (numeric or $\NIL$)}\\
       (\CONS\,\, x_1\,\, x_2) & \mbox{\em --- constructor} \\
       (\CAR\,\, x) & \\
       (\CDR\,\, x) & \mbox{\em --- selectors} \\
       (\NULLQ\,\, x) & \\
       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester/generic arithmetic} \\
%%       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
       (f\,\, x_1\,\,\ldots\,\, x_n) & \mbox{\em --- function application}
    \end{array}\right.
\end{eqnarray*}
%\vspace*{-6pt}
  \caption{The syntax of our language}\label{fig:lang-syntax}
%\figrule
\normalsize
%\vspace*{-6pt}
\end{figure}

@@

Closure  variables  get  their  liveness values  through  a  chain  of
dependences beginning  at a  variable at an  evaluation point.   As an
example,  consider again  the program  in Figure~\ref{fig:lv-closure}.
The variable  \pz\ returned  at $\psi_3$ depends  on \py\  through the
expression $(+~\py~1)$.  $\py$ in turn depends on the closure variable
$\px_{\mathit{\pi}_1}$  through $(\length~\px_{\mathit{\pi}_1})$.   We
denote this  chain of  dependences as  $\lbrack \psi_3\!\!\!:\!\!\!\pz
\leftarrow            (+~\py~1),\,           \py            \leftarrow
(\length~\px_{\mathit{\pi}_1})\rbrack$.    Indeed,   the   chains   of
closures  in   the  heap  are   a  runtime  representation   of  these
dependences.   Since  $\pz$  is  evaluated  at  $\psi_3$  due  to  the
expression $\SRETURN~\pz$,  the demand made by  the calling context(s)
of  $f$ places  a demand  on  $\pz$ which  will impart  a liveness  to
$\px_{\pi_1}$.  Other dependence chains which result in a liveness for
$\px_{\pi_1}$    are    $\lbrack    \psi_1\!\!\!:\!\!\py    \leftarrow
(\length~\px_{\pi_1})\rbrack$   and  $\lbrack   \psi_2\!\!\!:\!\!\!\pw
\leftarrow          (/~\pu~\py),\,\,           \py          \leftarrow
(\length~\px_{\pi_1})\rbrack$.   The  liveness analysis  described  in
this section declares  the liveness of $\px_{\pi_1}$ to be  a union of
the liveness arising  from these dependence chains.  To be  safe, a GC
during evaluation  of $\py$ at  $\psi_1$ has  to use this  liveness to
copy the heap starting from  $\px_{\pi_1}$.  However, notice that if a
GC  takes place  while evaluating  $\pz$  at $\psi_3$,  it can  safely
consider only the liveness arising from the dependence chain\linebreak
$\lbrack \psi_3\!\!\!:\!\!\!\pz \leftarrow (+~\py~1),\, \py \leftarrow
(\length~\px_{\pi_1})\rbrack$.    The    garbage   collection   scheme
described in Section~\ref{sec:GC-scheme} uses a generalization of this
observation  to  dynamically  select   an  evaluation  point  specific
liveness in order to collect more garbage.

@@

%\vspace*{-6.5mm}
\section{The GC scheme}
\label{sec:GC-scheme}

Our experimental setup  consists of an interpreter
for  our  language,  a liveness  analyzer,  and  a
single generation copying  collector.  The garbage
collector can be configured to work on the
basis  of  reachability  (RGC  mode)  or  use  
liveness  DFAs  (LGC  mode).

In        the       liveness        analysis       described        in
Section~\ref{sec:liveness-analysis},   the  liveness   of  a   closure
variable is  derived from dependence  chains along all paths  from the
program point  where the closure  was created to  reachable evaluation
points.  Assume,  for the  sake of  concreteness, that  $\epath_1$ and
$\epath_2$ are two such evaluation points.  During GC,
we would like to use a more precise liveness, based on the actual paths
taken  during  execution.   Therefore   we  create  separate  liveness
automata for dependences  along paths to $\epath_1$  and $\epath_2$, in
addition to  automata for dependences  along paths to  both $\epath_1$
and $\epath_2$.  The closure carries  the liveness environment for its
free  variables (as  pointers  to automata,  one  for each  variable).
Initially the liveness  environment is based on  the dependences along
both  $\epath_1$   and  $\epath_2$.   However,  after   evaluating  an
\SIF\ condition, the liveness environments are updated to one based on
either   $\epath_1$  or   $\epath_2$,  so   that  subsequent   garbage
collections are based on more precise liveness information.

Based on  the above considerations,  we restrict the  possible garbage
collection points in a function body to the following:
\begin{enumerate}
\item We  maintain static estimates  of the memory required  to create
  the closures for each function body.  On entering a function, if the
  available memory is  less than this requirement for  the function, a
  GC is triggered.
\item Since the evaluation of a  \SIF\ condition may trigger a 
  collection, after evaluating the condition   the available memory
  is  checked once  again against  a  revised estimate  of the  memory
  (based on  value of the condition)  required to execute the  rest of
  the program.  A GC is triggered if enough memory is unavailable.
\end{enumerate}

@@

%\subsection{Garbage collection for references on print stack}
The evaluation of  the top-level expression in a program  is driven by
a printing function~(Section
\ref{sec:semantics}).  
%% Printing of  atomic values is simple  and does
%% not  trigger a  GC.   In case  of  \CONS\ cells,  the
%% \CAR\ is  first printed followed  by \CDR.   In a lazy  language, both
%% \CAR\ and \CDR\  could be closures requiring evaluation,  and this may
%% trigger  more GCs.  When  this happens.   we have  to
%% consider any  references that  might be  on the  print stack  and copy
%% them.
We extend liveness-based  garbage  collection  to
this function.
\section{Experimental evaluation}
\label{sec:experiments}
Our benchmark  consists of programs taken  from nofib~\cite{nofib} and
other repositories for functional programs~\cite{PLT-Scheme, gc_bench,
  huffman-sicp}.     We    ran    the   experiments    on    8    core
Intel\textsuperscript{\textregistered}        Core\textsuperscript{TM}
i7-4770 3.40GHz CPU  having 8192KB L2 cache, 16GB RAM,  running 64 bit
Ubuntu 14.04.

@@

In addition to  evaluated data, our collector  also reclaims closures.
For this, we  had to modify the standard run-time  representation of a
closures  to carry  liveness  of its free  variables and  periodically
update the liveness during execution  time.  As expected, our garbage
collector  reclaims more  garbage and  reduces memory  requirement of
programs.  An  additional benefit is  that in  spite of using  a more
expensive collector, the execution times remains comparable in
most cases, and even improves for some programs.

>>>>

\documentclass[9pt,preprint,nonatbib]{sigplanconf}
\usepackage{balance}
\usepackage{float}
\usepackage{mathrsfs}
%%\usepackage{lineno}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{etex}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsmath}
%\usepackage[pdf]{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
%\usepackage{pst-rel-points}
\usepackage{bcprules}
\usepackage{subfigure}
\usepackage[boxed]{algorithm2e}
\usepackage{makecell}
\usepackage[numbers,sort&compress]{natbib}
% to save space in lists
\usepackage{enumitem}
\setlist{nolistsep}

@@

% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e., three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.
%\cmt{{
%
%\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors
%section.
%
%\author{
% 1st. author
\authorinfo {Prasanna Kumar. K} 
       {IIT Bombay, Mumbai 400076, India}
       {prasannak@cse.iitb.ac.in}
% 2nd. author
\authorinfo {Amitabha Sanyal} 
       {IIT Bombay, Mumbai 400076, India}
       {as@cse.iitb.ac.in}
% 3rd. author
\authorinfo {Amey Karkare}
       {IIT Kanpur, Kanpur 208016, India}
       {karkare@cse.iitk.ac.in}

@@

While  the runtime  system  of most  functional  languages includes  a
garbage collector to efficiently  reclaim memory, empirical studies on
Scheme~\cite{karkare06effectiveness}   and  on
Haskell~\cite{rojemo96lag} programs have shown that garbage collectors
leave uncollected a large number  of memory objects that are reachable
but not live (here {\em live} means the object can potentially be used
by the program at a  later stage).  This results in unnecessary memory
retention.

@@

\added{While  the idea  of  using static  analysis  to improve  memory
  utilization  has  been  shown  to   be  effective  for  {\em  eager}
  languages~\cite{asati14lgc,       HofmannJ03,       inoue88analysis,
    lee05static}, a straightforward extension  of the technique is not
  possible  for  lazy  languages,  where  heap-allocated  objects  may
  include   closures    (runtime   representations    of   unevaluated
  expressions).   }   \todelete{Designing  a  liveness  based  garbage
  collector   for  lazy   languages  poses   significant  challenges.}
Firstly, since  data is made  live by  evaluation of closures,  and in
lazy languages  the place in  the program where this  evaluation takes
place cannot  be statically determined, laziness  complicates liveness
analysis itself.  Moreover, for liveness-based  GC to be effective, we
need to  extend it  to closures  apart from  evaluated data.   Since a
closure can escape the scope in which  it was created, it has to carry
the  liveness  information  of   its  free  variables.   As  execution
progresses  and possible  future uses  are eliminated,  we update  the
liveness  information  in  a  closure with  a  more  precise  version.
For  these  reasons,  the  garbage   collector  also  becomes
significantly more  complicated than a liveness-based  collector for a
eager language.

@@

  Figure~\ref{fig:mot-example}   shows  an   example
program and the state of the heap at the program point $\pi$\footnote{We write
$\pi\!:\!e$ to  associate a  label $\pi$ with  the program  point just
before  the  expression  $e$. The label is not part of the syntax.},  i.e., just  before  the  evaluation  of
$(\length\ \pz)$.  The heap is represented  by a graph in which a node
either  represents  atomic  values   ($\NIL$,  integers,  etc.),  or  a
\CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
(represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
  references} and represent values of variables or fields.  The figure
shows the  lists \px\  and \pz\  partially evaluated  due to  the
\SIF\ condition 
(\NULLQ~(\CAR~\pz)).
The edges  shown by thick  arrows are those  which are live  at $\pi$.
%% Only those  cell which have a future use  should be preserved
%% during garbage collection, all other
%% cells can be  reclaimed.
Thus if a GC takes  place at $\pi$ with the heap shown
in Figure~\ref{fig:mot-example}(b),  a liveness-based  collector (LGC)
will preserve only  the cells referenced by $\pz$, and  the live cells
constituting the  closure referenced by $(\CDR~\pz)$.   In contrast, a
reachability-based collector (RGC) will preserve all cells.

In  this work  we show  that  static analysis  of heap  data can  help
garbage  collectors   in  reclaiming   more  garbage.    The  specific
contributions of this paper are:
\added{
\begin{itemize}
\item We  propose an interprocedural liveness  based GC
  scheme  for a  lazy first-order functional  language and  prove its
  correctness.  To the  best of our knowledge, this is  the first work
  that uses  the results  of an  interprocedural liveness  analysis to
  garbage    collect    both     evaluated    data    and    closures.
  Thomas~\cite{Thomas19951} describes a  copying garbage collector for
  the Three  Instruction Machine (TIM)~\cite{Fairbairn1987}  that only
  preserves live closures  in a function's environment  (also called a
  frame). However, in the absence of  details, it is not clear whether
  a) the  scope of the  method is  interprocedural, and b)  it handles
  algebraic datatypes like lists (the original design of TIM did not).
  All  other  previous attempts~\cite{shaham01heap,  ran.shaham-sas03,
    shaham02estimating,  asati14lgc, karkare06effectiveness}  involved
  either imperative or eager functional languages.

@@

\begin{figure}[t]\footnotesize
\renewcommand{\arraystretch}{0.9}
\begin{eqnarray*}
   p \in \mathit{Prog} & \!\!\!::=\!\!\! & d_1 \ldots d_n \,\,\,\, e_\mainpgm
   \hspace{5em} \,\,\,\,\,\,\,\,\,\; \mbox{\em --- program}\\
    \mathit{df} \in Fdef & \!\!\!::=\!\!\! & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots
\,\,x_n)\,\,
    e)
    \hspace{0.2em} \ \ \ \ \ \ \ \ \  \mbox{\em --- function def} \\
e \in \mathit{Expr} & \!\!\!::=\!\!\! &
\left\{\begin{array}{@{}ll@{\hspace{1em}}l}
       (\SIF\,\, x\,\, e_1\,\, e_2) && \!\!\!\mbox{\em --- conditional} \\
       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) &&\!\!\! \mbox{\em --- let
binding} \\
       (\SRETURN\,\, x) && \!\!\!\mbox{\em --- return from function}
    \end{array}\right. \\
s \in \mathit{App} & \!\!\!::=\!\!\!  &
\left\{\begin{array}{@{}l@{\hspace{1.2em}}l}
       k & \mbox{\em --- constant (numeric or $\NIL$)}\\
       (\CONS\,\, x_1\,\, x_2) & \mbox{\em --- constructor} \\
       (\CAR\,\, x) &  \mbox{\em --- selects $1^{st}$ part of \CONS} \\
       (\CDR\,\, x) &  \mbox{\em --- selects $2^{nd}$ part of \CONS} \\
       (\NULLQ\,\, x) & \mbox{\em --- returns 0 if x is not $\NIL$} \\
       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em generic arithmetic} \\
%%       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
       (f\,\, x_1\,\,\ldots\,\, x_n) & \mbox{\em --- function application}
    \end{array}\right.
\end{eqnarray*}
%\vspace*{-6pt}
  \caption{The syntax of our language}\label{fig:lang-syntax}
%\figrule
\normalsize
%\vspace*{-6pt}
\end{figure}

@@

Closure  variables  get  their  liveness values  through  a  chain  of
dependences beginning  at a  variable at an  evaluation point.   As an
example,  in  Figure~\ref{fig:lv-closure},   a  dependence  chain  for
$\px_{\mathit{\pi}_1}$  begins   with  the   variable  $\pz$   at  the
evaluation  point $\psi_3$.  The  variable \pz\  returned at  $\psi_3$
depends on  \py\ through  the expression  $(+~\py~1)$.  $\py$  in turn
depends  on   the  closure  variable   $\px_{\mathit{\pi}_1}$  through
$(\length~\px_{\mathit{\pi}_1})$.  We denote this chain of dependences
as   $\lbrack  \psi_3\!\!\!:\!\!\!\pz   \leftarrow  (+~\py~1),\,   \py
\leftarrow (\length~\px_{\mathit{\pi}_1})\rbrack$.  Indeed, the chains
of  closures  in  the  heap  are a  runtime  representation  of  these
dependences.   Since  $\pz$  is  evaluated  at  $\psi_3$  due  to  the
expression $\SRETURN~\pz$,  the demand made by  the calling context(s)
of  $f$ places  a demand  on  $\pz$ which  will impart  a liveness  to
$\px_{\pi_1}$.  Other dependence chains which result in a liveness for
$\px_{\pi_1}$    are    $\lbrack    \psi_1\!\!\!:\!\!\py    \leftarrow
(\length~\px_{\pi_1})\rbrack$   and  $\lbrack   \psi_2\!\!\!:\!\!\!\pw
\leftarrow          (/~\pu~\py),\,\,           \py          \leftarrow
(\length~\px_{\pi_1})\rbrack$.   The  liveness analysis  described  in
this section declares  the liveness of $\px_{\pi_1}$ to be  a union of
the liveness arising  from these dependence chains.  To be  safe, a GC
during evaluation  of $\py$ at  $\psi_1$ has  to use this  liveness to
copy the heap starting from  $\px_{\pi_1}$.  However, notice that if a
GC  takes place  while evaluating  $\pz$  at $\psi_3$,  it can  safely
consider only the liveness arising from the dependence chain\linebreak
$\lbrack \psi_3\!\!\!:\!\!\!\pz \leftarrow (+~\py~1),\, \py \leftarrow
(\length~\px_{\pi_1})\rbrack$.    The    garbage   collection   scheme
described in Section~\ref{sec:GC-scheme} uses a generalization of this
observation  to  dynamically  select   an  evaluation  point  specific
liveness in order to collect more garbage.

@@

%\vspace*{-6.5mm}
\section{The GC scheme}
\label{sec:GC-scheme}
In        the       liveness        analysis       described        in
Section~\ref{sec:liveness-analysis},   the  liveness   of  a   closure
variable is  derived from dependence  chains along all paths  from the
program point  where the closure  was created to  reachable evaluation
points.  Assume,  for the  sake of  concreteness, that  $\epath_1$ and
$\epath_2$ are two such evaluation points.  During GC,
we would like to use a more precise liveness, based on the actual paths
taken  during  execution.   Therefore   we  create  separate  liveness
automata for dependences  along paths to $\epath_1$  and $\epath_2$, in
addition to  automata for dependences  along paths to  both $\epath_1$
and $\epath_2$.  The closure carries  the liveness environment for its
free  variables (as  pointers  to automata,  one  for each  variable).
Initially the liveness  environment is based on  the dependences along
both  $\epath_1$   and  $\epath_2$.   However,  after   evaluating  an
\SIF\ condition, the liveness environments are updated to one based on
either   $\epath_1$  or   $\epath_2$,  so   that  subsequent   garbage
collections are based on more precise liveness information.

Based on  the above considerations,  we restrict the  possible garbage
collection points in a function body to the following:
\begin{enumerate}
\item We  statically over-approximate the memory required  to create
  the closures for each function body. On entering a function, if the
  available memory is  less than this requirement for  the function, a
  GC is triggered.
\item Since the evaluation of a  \SIF\ condition may trigger a 
  collection, after evaluating the condition   the available memory
  is  checked once  again against  a  revised estimate  of the  memory
  (based on  value of the condition)  required to execute the  rest of
  the program.  A GC is triggered if enough memory is unavailable.
\end{enumerate}

@@

%\subsection{Garbage collection for references on print stack}
The evaluation of  the top-level expression in a program  is driven by
a printing function~(Section
\ref{sec:semantics}).  
%% Printing of  atomic values is simple  and does
%% not  trigger a  GC.   In case  of  \CONS\ cells,  the
%% \CAR\ is  first printed followed  by \CDR.   In a lazy  language, both
%% \CAR\ and \CDR\  could be closures requiring evaluation,  and this may
%% trigger  more GCs.  When  this happens.   we have  to
%% consider any  references that  might be  on the  print stack  and copy
%% them.
We extend liveness-based  garbage  collection  to
this function.
\section{Experimental evaluation}
\label{sec:experiments}
Our experimental setup  consists of an interpreter
for  our  language,  a liveness  analyzer,  and  a
single generation copying  collector.  The garbage
collector can be configured to work on the
basis  of  reachability  (RGC  mode)  or  use  
liveness  DFAs  (LGC  mode).

Our benchmark  consists of programs taken  from nofib~\cite{nofib} and
other repositories for functional programs~\cite{PLT-Scheme, gc_bench,
  huffman-sicp}.     We    ran    the   experiments    on    8    core
Intel\textsuperscript{\textregistered}        Core\textsuperscript{TM}
i7-4770 3.40GHz CPU  having 8192KB L2 cache, 16GB RAM,  running 64 bit
Ubuntu 14.04.

@@

In addition to  evaluated data, our collector  also reclaims closures.
For this, we  had to modify the standard run-time  representation of
closures  to carry  liveness  of its free  variables and  periodically
update the liveness during execution  time.  As expected, our garbage
collector  reclaims more  garbage and  reduces memory  requirement of
programs.  An  additional benefit is  that in  spite of using  a more
expensive collector, the execution times remains comparable in
most cases, and even improves for some programs.
